---
title: "Reproducing the Website Analyses"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the Website Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(salem)
```

The [Salem Witchcraft Website](https://www.tulane.edu/~salem/index.html)
presents a number of data analyses and explains how to reproduce them
using Excel. This vignette will document how to as closely as possible
reproduce those analyses using R.  In some cases the analyses will be 
extended in order to illustrate points in the website text that were
not easily expressible in Excel. 

While the website has seven data sets, the package has only 5. This is 
because R makes it relatively simple to manipulate the data and also
because this organization is more in line with how social scientists would 
likely create a data set today.

As is always true in R, there are many different ways to do the same thing. 
If you are using this for teaching purposes you will likely want to use
the approach that most closely matches what you are doing in your course and
what your goals are for students to work on their own analyses. The focus
of this document is on use of _ggplot2_ for graphing and in a subsequent iteration
will also illustrate use of the base R plotting system. There are some 
examples that use the tidyverse for analysis and some that use base R. 
However, even within these major groupings there are many ways to acheive
the same goals.  

Most of the analyses presented are graphical and so that will be the focus 
here. However some preprocessing will also be illustrated.

This version does not yet include spatial analysis, but that is on the.

For the ggplots this theme is being used in may cases.

```{r}
salem_theme <-  theme(axis.text.x = element_text(angle = 45, vjust = .5))
```


## Ggplot and the Tidyverse

### Accusations.

The website includes displays of the distribution by month.

#### Histogram and frequency table by month

```{r fig.width=5}
ggplot(data = accused_witches) + 
       aes(x = Month.of.Accusation.Name) +
       geom_bar() +
       #geom_dotplot(dotsize = .17, method = "histodot", binwidth = 1, fill = "red") +
       labs(title = "Accusations by month", x = "Month")  +
       scale_x_discrete(drop = FALSE) +
       salem_theme

table(accused_witches$Month.of.Accusation.Name, useNA = "ifany", 
      dnn = "Month") %>% kable(caption = "Accusations by Month")

```



#### Cumulative accusations by month
```{r fig.width=5}
ggplot(data = accused_witches) + 
       aes(x = as.numeric(Month.of.Accusation.Name)) +
       stat_ecdf(na.rm = TRUE, geom = "line") +
       labs(title = "Cumulative accusations by month", x = "Month", 
            y = "Cumulative Accusations")  +
       scale_x_continuous(breaks = c(1:12), 
                          labels = c("Jan", "Feb", "Mar", "Apr", "May", 
                                     "Jun", "Jul","Aug", "Sept", "Oct", "Nov", "Dec") )


cumsum(table(accused_witches$Month.of.Accusation.Name)) %>% kable()

```

An alternative approach using dplyr that provides some other options. This approach
would also let you add other variables to the new monthly data, whether the 
number of executions or other data you collect. One advantage of this approach
is that it is easier to create a Pareto Histogram as displayed on the 
website. These examples show three possible ways to display the Pareto 
histogram.  As usual there are other approaches. 

```{r message=FALSE, fig.width=5}
monthly_accusations <- accused_witches %>% 
                      filter(!is.na(Month.of.Accusation)) %>%
                      group_by(Month.of.Accusation.Name) %>%
                      summarize(number = n()) %>% 
                      mutate(cumulative = cumsum(number) )
monthly_accusations %>% kable(caption = "Accusations by Month")

ggplot(data = monthly_accusations) +
             aes(x = Month.of.Accusation.Name, y = cumulative, group =1) +
             geom_col(aes(y = number)) +
             geom_line() +
             labs(title = "Frequency and Cumulative Frequency of Accusations by Month") +
             salem_theme


 monthly_accusations %>% 
             arrange(desc(number))  %>% 
            mutate(Month.of.Accusation.Name = 
                   factor(Month.of.Accusation.Name, 
                         levels=Month.of.Accusation.Name)) %>%
            mutate(pareto_cumulative = cumsum(number) ) %>%
        ggplot() +
             aes(x = Month.of.Accusation.Name, y = number) +
             geom_col() +
             labs(title = "Pareto Histogram of Accusations by Month") +
             salem_theme


monthly_accusations %>% 
             arrange(desc(number))  %>% 
            mutate(Month.of.Accusation.Name = 
                   factor(Month.of.Accusation.Name, 
                         levels=Month.of.Accusation.Name)) %>%
            mutate(pareto_cumulative = cumsum(number) ) %>%
        ggplot() +
             aes(x = Month.of.Accusation.Name, y = pareto_cumulative) +
             geom_point( ) +
             geom_path(aes(y=pareto_cumulative, group=1), colour="slateblue1", lty=3, size=0.9) +
             labs(title = "Pareto Cumulative Distribution of Accusations by Month")+
             salem_theme


 monthly_accusations %>% 
             arrange(desc(number))  %>% 
            mutate(Month.of.Accusation.Name = 
                   factor(Month.of.Accusation.Name, 
                         levels=Month.of.Accusation.Name)) %>%
            mutate(pareto_cumulative = cumsum(number) ) %>%
        ggplot() +
             aes(x = Month.of.Accusation.Name, y = number) +
             geom_col() +
             geom_path(aes(y=pareto_cumulative, group=1), colour="slateblue1", lty=3, size=0.9) +
             labs(title = "Pareto Histogram of Accusations by Month") +
             salem_theme


```

#### Chronology of Executions

Creating the bar chart of executions over time is similar to creating
that for accusations.  Additionally, these are shown as what is 
termed scatterplots but now would be called line graphs.  

Since most accused were not executed (and this is indicated by setting them
to NA or missing) we'll remove the missing cases from the graph.

```{r fig.width=5}
ggplot(data = na.omit(accused_witches)) + 
       aes(x = Month.of.Execution.Name) +
       geom_bar() +
       #geom_dotplot(dotsize = .17, method = "histodot", binwidth = 1, fill = "red") +
       labs(title = "Executions by month", x = "Month")  +
       scale_x_discrete(drop = FALSE) +
       salem_theme


table(accused_witches$Month.of.Execution.Name, useNA = "ifany", 
      dnn = "Month") %>% kable(caption = "Executions by Month")

```

There are many ways to explore the relationship between executions and 
accusations.  On the website this is displayed using overlapping 
line plots of the number of execuations in a month and the number of 
accusations.  If teaching, it is important to point out that the execuations
in a month were not generally of the same people who were accused in that 
month.  To reproduce this, one approach is to again use summarize() to create
a monthly data set.

```{r fig.width=6}

monthly_accused <- accused_witches %>% 
                      filter(!is.na(Month.of.Accusation)) %>%
                      group_by(Month.of.Accusation.Name) %>%
                      summarize(accusations = n())  %>%
                      rename(Month = Month.of.Accusation.Name)
monthly_executed <- accused_witches %>% 
                      filter(!is.na(Month.of.Execution)) %>%
                      group_by(Month.of.Execution.Name) %>%
                      summarize(executions = n()) %>%
                      rename(Month = Month.of.Execution.Name)
monthly_data <- left_join(monthly_accused, monthly_executed) %>%
                          mutate(executions = 
                                   ifelse(is.na(executions), 0, executions))


ggplot(monthly_data) +
     aes(x = Month, y = accusations, group =1) +
     geom_path(color = "blue" ) +
     geom_line(aes(y = executions, x = Month, group=1, color = "red")) +
     labs(title = "Executions and Accusations by Month") +
     salem_theme
```




### Geographic Spread of Accusations

The pivot table of month of accusation by town. However that table is huge
and only the first seven columns are shown on the website. A more compact 
display is given below rather than using kable().  Also note that 
the row and column totals are added. 

```{r}
addmargins(table( accused_witches$Month.of.Accusation.Name, accused_witches$Residence,
        dnn = c("Month", "Residence") )) 
```

The site discusses the "grand total" data (the row sum) as reflecting the
varying intensity in the different towns.  This can also be shown using
a univariate table (or dplyr::summarize())

```{r}
table(accused_witches$Residence, dnn = "Residence") %>% 
  kable(caption = "Residences of Accused")

accused_witches %>% group_by(Residence) %>% summarize(number = n()) %>% 
     kable(caption = "Residences of Accused")

```

The website contains a towns data set that essentially reproduces the large
contingency table.  This makes sense to do in Excel, but in R we do not 
need to makea separate data set. 

To make a histogram for each town we can select the town we want. This is 
illustrated below for Andover.

Note that in the data set the white space from the beginning and ending of the
place names has been removed. 


```{r fig.width=5}

accused_witches %>% filter(Residence == "Andover") %>%
   ggplot() +
   aes(x = Month.of.Accusation.Name) +
   geom_bar() +
   scale_x_discrete(drop=FALSE) +
   salem_theme


```

The website encourages examining the period before and after the convening of the ...
in June.  To help make this more visible we can add a vline.

```{r fig.width=5}

accused_witches %>% filter(Residence == "Andover") %>%
   ggplot() +
   aes(x = Month.of.Accusation.Name) +
   geom_bar() +
   geom_vline(xintercept = 6.5, color = "red") +
   scale_x_discrete(drop=FALSE) +
   salem_theme

```

Alternatively, we can use faceting to show all of the histograms. 

```{r fig.width=6, fig.height=8}

accused_witches %>% 
   ggplot() +
   aes(x = Month.of.Accusation.Name) +
   geom_bar() +
   geom_vline(xintercept = 6.5, color = "red") +
   scale_x_discrete(drop = FALSE) +
   facet_wrap(facets = vars(Residence), ncol = 4) +
   labs(title = "Number of accusations by month",
        subtitle = "Red line indicates reconvening of the Court of Oyer and Terminer",
        x = "Month of Accusation") +
   theme(axis.text.x = element_text(angle = 90, vjust = .5))

```

## Social Conflict

The website explores a variety of dimensions of social conflict in the Salem
region.  One of the important social cleavages was between those who 
did or did not support Reverend Parris, who had been appointed as
the minister for the new church in Salem Village. (This was the Puritan
church, which was the official church in Massacheusetts.)  

The site has two data sets, one of pro-Parris individuals and one of 
anti-Parris individuals.  For use in R these have been combined with 
an added variable identifying the position of the indivual. This data set is
called parris_social. Also the original column label for the anti-Parris
data was "[Church] Member" while for the pro-Parris data it was "Church-Member."
In the parris_social data Church-Member is used. 
Using the variable `view` it can be subset to pro- and
anti- groups.  However it is not necessary to do this.

A three-way table can be created, as one example. Two ways of doing this
are shown.  Marginals and proportions could be added.


```{r}

table(parris_social$Sex, parris_social$Identification, parris_social$view,
      dnn = c("Sex", "Identification", "View"))

ftable(parris_social$Sex, parris_social$Identification, parris_social$view,
      dnn = c("Sex", "Identification", "View"))


```

Treating position on Parris as the dependent varable, we might rearrange the tables.

```{r}

table(parris_social$view, parris_social$Identification, parris_social$Sex, 
      dnn = c("View", "Identification", "Sex"))

ftable( parris_social$Identification, parris_social$view,  parris_social$Sex,
      dnn = c("View", "Identification", "Sex"))


```


The parris_social data set focuses on those who signed pro- and anti-Parris 
petitions.  In contrast, the salem_village data set includes individuals who did not 
sign petitions in addition to those who did.  However, it is limited to those who paid
taxes so many of the young men and women of all ages who are in the parris_social data
are not included.  This also includes data from multiple years.  

The basic table for this is the analysis of church membership and position on Parris.

```{r}
table(salem_village$Petition, salem_village$Church.to.1696, 
      dnn = c( "Membership", "View"))

addmargins( table(salem_village$Petition, salem_village$Church.to.1696, 
      dnn = c( "Membership", "View")))
```

#### Political Power

As explained on the website, the Village of Salem was governed by a Committee of 
Five in addition to the Town Meeting 
[see Salem Village Politics](https://www.tulane.edu/~salem/Politics%20Intro.html).

The data set that is part of this package called committee_list is 
structured differently than the original data. It is in "long" format with
one row per person-year for each committee member for the years they 
served.   This can be coverted to various other formats if desired.  

You can make a table see how the make up of the committee changed over the period
from 1687 to 1698.

```{r}

table(committee_list$Petition, committee_list$Year, dnn = c("Petition", "Year"))

```

The website shows a complex pivot table involving three variables.

```{r}
table( committee_list$Petition, committee_list$Social, committee_list$Year )
```

One way to explore the association is to cross tabulate the Petition and Social
variables.  However, this counts the same people multiple times if they served
for multiple years.

```{r}

table( committee_list$Petition, committee_list$Social )

```

To adjust for this and get the data matching the website
we need to remove those duplicates. Then the table matches that found at
[Committee Social Report](https://www.tulane.edu/~salem/Committeemen%20Social%20Report.html).

```{r}
committee_list2 <-  committee_list %>% group_by(Committee.Members) %>% 
                             summarize(Social = first(Social), 
                                       Petition = first(Petition),
                                       Terms = n())

addmargins(table( committee_list2$Petition, committee_list2$Social ))
```


### Wealth and Salem Village Conflict

The tax_comparison data set is also set up in long format. This can then be
used to create a comparison table similar to the table on 
[Tax Comparison](https://www.tulane.edu/~salem/1690%20Tax%20Comparisons.html).
One limitation is that R does not have a built in function for the mode.

```{r}
tax_comparison %>% group_by(Year, Petition) %>%
            summarize(mean = mean(Tax), median = median(Tax), minimum = min(Tax),
                      maximum =  max(Tax), range = maximum - minimum, 
                      sum = sum(Tax), Count = n())
```

## Conclusion

This has illustrated how to reproduce the analyses from the Salem Witchcraft 
Website. The raw data is available at the websie and at 
[the respoistory for this package](https://github.com/elinw/salem/tree/master/rawdata).
The code used to transform the original data into the files presented here is
also available at that location.
