---
title: "Reproducing the Website Analyses"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the Website Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(salem)
```

The [Salem Witchcraft Website](https://www.tulane.edu/~salem/index.html)
presents a number of data analyses and explains how to reproduce them
using Excel. This vignette will document how to as closely as possible
reproduce those analyses using R.  In some cases the analyses will be 
extended in order to illustrate points in the website text that were
not easily expressible in Excel. 

While the website has seven data sets, the package has only 5. This is 
because R makes it relatively simple to manipulate the data and also
because this organization is more in line with how social scientists would 
likely create a data set today.

As is always true in R, there are many different ways to do the same thing. 
If you are using this for teaching purposes you will likely want to use
the approach that most closely matches what you are doing in your course and
what your goals are for students to work on their own analyses. The focus
of this document is on use of _ggplot2_ for graphing and in subsequent iterations
will also illustrate use of the base R plotting system. There are some 
examples that use the tidyverse for analysis and some that use base R. 
However, even within these major groupings there are many ways to acheive
the same goals. We have aimed to present ways to do this that 
are relatively accessible to people who have not used R extensively. 

Many of the analyses presented are graphical and so that will be the focus 
here. However some preprocessing will also be illustrated.

This version does not yet include spatial analysis, but that is on the roadmap.

For the ggplots this theme is being used in may cases.

```{r}
salem_theme <-  theme(axis.text.x = element_text(angle = 45, vjust = .5))
```


## Ggplot and the Tidyverse

### Accusations

The website includes displays of the distribution by month.

#### Histogram and frequency table by month

```{r fig.width=5}
ggplot(data = accused_witches) + 
       aes(x = Month.of.Accusation.Name) +
       geom_bar() +
       #geom_dotplot(dotsize = .17, method = "histodot", binwidth = 1, fill = "red") +
       labs(title = "Accusations by month", x = "Month")  +
       scale_x_discrete(drop = FALSE) +
       salem_theme

table(accused_witches$Month.of.Accusation.Name, useNA = "ifany", 
      dnn = "Month") %>% kable(caption = "Accusations by Month")

```



#### Cumulative accusations by month
```{r fig.width=5}
ggplot(data = accused_witches) + 
       aes(x = as.numeric(Month.of.Accusation.Name)) +
       stat_ecdf(na.rm = TRUE, geom = "line") +
       labs(title = "Cumulative accusations by month", x = "Month", 
            y = "Cumulative Accusations")  +
       scale_x_continuous(breaks = c(1:12), 
                          labels = c("Jan", "Feb", "Mar", "Apr", "May", 
                                     "Jun", "Jul","Aug", "Sept", "Oct", "Nov", "Dec") )


cumsum(table(accused_witches$Month.of.Accusation.Name)) %>% kable()

```

An alternative approach using dplyr that provides some other options. This approach
would also let you add other variables to the new monthly data, whether the 
number of executions or other data you collect. One advantage of this approach
is that it is easier to create a Pareto Histogram as displayed on the 
website. These examples show three possible ways to display the Pareto 
histogram.  As usual there are other approaches. 

```{r message=FALSE, fig.width=5}
monthly_accusations <- accused_witches %>% 
                      filter(!is.na(Month.of.Accusation)) %>%
                      group_by(Month.of.Accusation.Name) %>%
                      summarize(number = n()) %>% 
                      mutate(cumulative = cumsum(number) )
monthly_accusations %>% kable(caption = "Accusations by Month")

ggplot(data = monthly_accusations) +
             aes(x = Month.of.Accusation.Name, y = cumulative, group =1) +
             geom_col(aes(y = number)) +
             geom_line() +
             labs(title = "Frequency and Cumulative Frequency of Accusations by Month") +
             salem_theme


 monthly_accusations %>% 
             arrange(desc(number))  %>% 
            mutate(Month.of.Accusation.Name = 
                   factor(Month.of.Accusation.Name, 
                         levels=Month.of.Accusation.Name)) %>%
            mutate(pareto_cumulative = cumsum(number) ) %>%
        ggplot() +
             aes(x = Month.of.Accusation.Name, y = number) +
             geom_col() +
             labs(title = "Pareto Histogram of Accusations by Month") +
             salem_theme


monthly_accusations %>% 
             arrange(desc(number))  %>% 
            mutate(Month.of.Accusation.Name = 
                   factor(Month.of.Accusation.Name, 
                         levels=Month.of.Accusation.Name)) %>%
            mutate(pareto_cumulative = cumsum(number) ) %>%
        ggplot() +
             aes(x = Month.of.Accusation.Name, y = pareto_cumulative) +
             geom_point( ) +
             geom_path(aes(y=pareto_cumulative, group=1), colour="slateblue1", lty=3, size=0.9) +
             labs(title = "Pareto Cumulative Distribution of Accusations by Month")+
             salem_theme


 monthly_accusations %>% 
             arrange(desc(number))  %>% 
            mutate(Month.of.Accusation.Name = 
                   factor(Month.of.Accusation.Name, 
                         levels=Month.of.Accusation.Name)) %>%
            mutate(pareto_cumulative = cumsum(number) ) %>%
        ggplot() +
             aes(x = Month.of.Accusation.Name, y = number) +
             geom_col() +
             geom_path(aes(y=pareto_cumulative, group=1), colour="slateblue1", lty=3, size=0.9) +
             labs(title = "Pareto Histogram of Accusations by Month") +
             salem_theme


```

#### Chronology of Executions

Creating the bar chart of executions over time is similar to creating
that for accusations.  Additionally, these are shown as what is 
termed scatterplots but now would be called line graphs.  

Since most accused were not executed (and this is indicated by setting them
to NA or missing) we'll remove the missing cases from the graph.

```{r fig.width=5}
ggplot(data = na.omit(accused_witches)) + 
       aes(x = Month.of.Execution.Name) +
       geom_bar() +
       #geom_dotplot(dotsize = .17, method = "histodot", binwidth = 1, fill = "red") +
       labs(title = "Executions by month", x = "Month")  +
       scale_x_discrete(drop = FALSE) +
       salem_theme


table(accused_witches$Month.of.Execution.Name, useNA = "ifany", 
      dnn = "Month") %>% kable(caption = "Executions by Month")

```

There are many ways to explore the relationship between executions and 
accusations.  On the website this is displayed using overlapping 
line plots of the number of executions in a month and the number of 
accusations.  If teaching, it is important to point out that the execuations
in a month were not generally of the same people who were accused in that 
month.  To reproduce this, one approach is to again use summarize() to create
a monthly data set.

```{r fig.width=6}

monthly_accused <- accused_witches %>% 
                      filter(!is.na(Month.of.Accusation)) %>%
                      group_by(Month.of.Accusation.Name) %>%
                      summarize(accusations = n())  %>%
                      rename(Month = Month.of.Accusation.Name)
monthly_executed <- accused_witches %>% 
                      filter(!is.na(Month.of.Execution)) %>%
                      group_by(Month.of.Execution.Name) %>%
                      summarize(executions = n()) %>%
                      rename(Month = Month.of.Execution.Name)
monthly_data <- left_join(monthly_accused, monthly_executed) %>%
                          mutate(executions = 
                                   ifelse(is.na(executions), 0, executions))


ggplot(monthly_data) +
     aes(x = Month, y = accusations, group = 1) +
     geom_path( color = "blue" ) +
     geom_line(aes(y = executions, x = Month, group = 1), color = "red") +
     labs(title = "Executions and Accusations by Month",
          subtitle = "Accusations in blue and executions in red") +
     salem_theme
```




### Geographic Spread of Accusations

The pivot table of month of accusation by town. However that table is huge
and only the first seven columns are shown on the website. A more compact 
display is given below rather than using kable().  Also note that 
the row and column totals are added. 

```{r}
addmargins(table( accused_witches$Month.of.Accusation.Name, accused_witches$Residence,
        dnn = c("Month", "Residence") )) 
```

The site discusses the "grand total" data (the row sum) as reflecting the
varying intensity in the different towns.  This can also be shown using
a univariate table (or dplyr::summarize())

```{r}
table(accused_witches$Residence, dnn = "Residence") %>% 
  kable(caption = "Residences of Accused")

accused_witches %>% group_by(Residence) %>% summarize(number = n()) %>% 
     kable(caption = "Residences of Accused")

```

The website contains a towns data set that essentially reproduces the large
contingency table.  This makes sense to do in Excel, but in R we do not 
need to makea separate data set. 

To make a histogram for each town we can select the town we want. This is 
illustrated below for Andover.

Note that in the data set the white space from the beginning and ending of the
place names has been removed. 


```{r fig.width=5}

accused_witches %>% filter(Residence == "Andover") %>%
   ggplot() +
   aes(x = Month.of.Accusation.Name) +
   geom_bar() +
   scale_x_discrete(drop=FALSE) +
   salem_theme


```

The website encourages examining the period before and after the convening of 
the Court of Oyer and Terminer in June.  
To help make this more visible we can add a vline.

```{r fig.width=5}

accused_witches %>% filter(Residence == "Andover") %>%
   ggplot() +
   aes(x = Month.of.Accusation.Name) +
   geom_bar() +
   geom_vline(xintercept = 6.5, color = "red") +
   scale_x_discrete(drop = FALSE) +
  labs(title = "Accusations by Month", 
       subtitle = "Red line indicates reconvening of the Court of Oyer and Terminer",
       x = "Month of accusation") +
   salem_theme

```

Alternatively, we can use faceting to show all of the histograms. 

```{r fig.width=6, fig.height=8}

accused_witches %>% 
   ggplot() +
   aes(x = Month.of.Accusation.Name) +
   geom_bar() +
   geom_vline(xintercept = 6.5, color = "red") +
   scale_x_discrete(drop = FALSE) +
   facet_wrap(facets = vars(Residence), ncol = 4) +
   labs(title = "Number of accusations by month",
        subtitle = "Red line indicates reconvening of the Court of Oyer and Terminer",
        x = "Month of Accusation") +
   theme(axis.text.x = element_text(angle = 90, vjust = .5))

```

### Maps

We can use maps to represent the geographic distribution over time 
spatially.  The salem_region data set is an sf (simple features) data 
set that provides geographic data for the towns in the three Massachuessets 
counties represented in our data.  Not all of the towns in these three counties
had accusations: the majority did not. The data also include the total
accused for each town and the monthly accused for the towns that ever had an 
accusation. However, these figures should _never_ be used outside of maps. 
This is because many towns have multiple records. This is especially true if 
they include islands, which is common in this region of the state. 
These are based on modern municipal boundaries which are close to
but not the same as the historical ones. There are many geographic analyses
that can be done. Below a comparison of the distribution of whether towns
had at least on accusation in each of four months
is given.

First, create variables indicating whether towns did or did not have 
an accusation in a month and remove the TOWN_LABELs for all the towns that never
had an accusation.

```{r}

newdata <- salem_region %>% mutate(February.Any = February > 0, March.Any = March > 0,
                                   April.Any = April > 0, May.Any = May > 0,
                                   June.Any = June > 0, July.Any = July > 0,
                                   August.Any = August > 0, 
                                   September.Any = September >0,
                                   October.Any = October > 0,
                                   November.Any = November > 0)

newdata$TOWN_LABEL <- ifelse(newdata$n_accused == 0, NA,  newdata$TOWN_LABEL)
```

```{r fig.width=8, fig.height=8}
p1 <- ggplot(newdata) 
p2 <- geom_sf_text(aes(label = TOWN_LABEL), color = "blue", size = 2, nudge_x = 5,
                nudge_y = 5, na.rm = TRUE) 
p3 <-       scale_fill_manual(values = c( "grey", "red"), na.value = "white") 

p1  + geom_sf(data = newdata,  aes(fill = February.Any), color = "black", size = .1) +
  p3 + p2 + labs(title = "Location of Accusations in February") 

p1 +  geom_sf(data = newdata,  aes(fill = July.Any), color = "black", size = .1) +
  p3 + p2 + labs(title = "Location of Accusations in July") 


p1 + geom_sf(data = newdata,  aes(fill = August.Any), color = "black", size = .1) +
  p3 + p2  + labs(title = "Location of Accusations in August") 

p1 + geom_sf(data = newdata,  aes(fill = November.Any), color = "black", size = .1) +
  p3 + p2 + labs(title = "Location of Accusations in November") 

```







## Social Conflict

The website explores a variety of dimensions of social conflict in the Salem
region.  One of the important social cleavages was between those who 
did or did not support Reverend Parris, who had been appointed as
the minister for the new church in Salem Village. (This was the Puritan
church, which was the official church in Massacheusetts.)  



The site has two data sets, one of pro-Parris individuals and one of 
anti-Parris individuals.  For use in R these have been combined with 
an added variable identifying the position of the indivual. This data set is
called parris_social. Also the original column label for the anti-Parris
data was "[Church] Member" while for the pro-Parris data it was "Church-Member."
In the parris_social data Church-Member is used. 
Using the variable `view` it can be subset to pro- and
anti- groups.  However it is not necessary to do this.

A three-way table can be created, as one example. Two ways of doing this
are shown.  Marginals and proportions could be added.


```{r}

table(parris_social$Sex, parris_social$Identification, parris_social$view,
      dnn = c("Sex", "Identification", "View"))

ftable(parris_social$Sex, parris_social$Identification, parris_social$view,
      dnn = c("Sex", "Identification", "View"))


```

Treating position on Parris as the dependent varable, we might rearrange the tables.

```{r}

table(parris_social$view, parris_social$Identification, parris_social$Sex, 
      dnn = c("View", "Identification", "Sex"))

ftable( parris_social$Identification, parris_social$view,  parris_social$Sex,
      dnn = c("View", "Identification", "Sex"))


```


The parris_social data set focuses on those who signed pro- and anti-Parris 
petitions.  In contrast, the salem_village data set includes individuals who did not 
sign petitions in addition to those who did.  However, it is limited to those who paid
taxes so many of the young men and women of all ages who are in the parris_social data
are not included.  This also includes data from multiple years.  

The basic table for this is the analysis of church membership and position on Parris.

```{r}
table(salem_village$Petition, salem_village$Church.to.1696, 
      dnn = c( "Membership", "View"))

addmargins( table(salem_village$Petition, salem_village$Church.to.1696, 
      dnn = c( "Membership", "View")))
```

#### Political Power

As explained on the website, the Village of Salem was governed by a Committee of 
Five in addition to the Town Meeting 
[see Salem Village Politics](https://www.tulane.edu/~salem/Politics%20Intro.html).

The data set that is part of this package called committee_list is 
structured differently than the original data. It is in "long" format with
one row per person-year for each committee member for the years they 
served.   This can be coverted to various other formats if desired.  

You can make a table see how the make up of the committee changed over the period
from 1687 to 1698.

```{r}

table(committee_list$Petition, committee_list$Year, dnn = c("Petition", "Year"))

```

The website shows a complex pivot table involving three variables.

```{r}
table( committee_list$Petition, committee_list$Social, committee_list$Year )
```

One way to explore the association is to cross tabulate the Petition and Social
variables.  However, this counts the same people multiple times if they served
for multiple years.

```{r}

table( committee_list$Petition, committee_list$Social )

```

To adjust for this and get the data matching the website
we need to remove those duplicates. Then the table matches that found at
[Committee Social Report](https://www.tulane.edu/~salem/Committeemen%20Social%20Report.html).

```{r}
committee_list2 <-  committee_list %>% group_by(Committee.Members) %>% 
                             summarize(Social = first(Social), 
                                       Petition = first(Petition),
                                       Terms = n())

addmargins(table( committee_list2$Petition, committee_list2$Social ))
```


### Wealth and Salem Village Conflict

The tax_comparison data set is also set up in long format. This can then be
used to create a comparison table similar to the table on 
[Tax Comparison](https://www.tulane.edu/~salem/1690%20Tax%20Comparisons.html).
One limitation is that R does not have a built in function for the mode.

This is the most complex analysis on the website. Complications include
people moving in and out of the data in different years (individuals appear
between one and six times), the changing 
amount of taxes over the time period and name changes. The analysis looks 
at both group and individual level comparisons and individual level mobility.


Here is the distribution of tax payers by year.

```{r}
table(tax_comparison$Year, dnn = c("Year")) %>% kable(caption = "Tax payers by year")
```

The next table summarizes descriptive statistics for all of the years. The
website general shows one year at a time or in some cases compares two years. 
You can add `filter(Year = 1681)`
to display only the 1681 data, changing the filter to select whatever years
you want. For example to select all of the data for 1681 through 1694 one way
would be to add `select(Year >= 1691 & Year <= 1694)`.   

Let's start by looking at the group level data.

The first table is for everyone in a year. The second separates the 
data based on which petition the tax payer signed (or if they) did not sign.

```{r}
tax_comparison %>% group_by(Year) %>%
            summarize(mean = round(mean(Tax), 1), median = round(median(Tax), 1),
                                   minimum = min(Tax),
                      maximum =  max(Tax), range = maximum - minimum, 
                      sum = sum(Tax), Count = n()) %>% 
                      kable(caption = "Taxes paid by year")



tax_comparison %>% group_by(Year, Petition) %>%
            summarize(mean = round(mean(Tax), 1), median = round(median(Tax), 1),
                                   minimum = min(Tax),
                      maximum =  max(Tax), range = maximum - minimum, 
                      sum = sum(Tax), Tax_payers = n()) %>% 
                      kable(caption = "Taxes paid by year and petition signed")
```

To do the analysis comparing the median tax paid for the three groups we can
graph the values over time.  In this case, we will first make a data set 
of statistics to work with.

```{r fig.width=5}
yearly_tax <- tax_comparison %>% group_by(Year, Petition) %>%
            summarize(mean = round(mean(Tax), 1), median = round(median(Tax), 1),
                                   minimum = min(Tax),
                      maximum =  max(Tax), range = maximum - minimum, 
                      sum = sum(Tax), Count = n())

ggplot(yearly_tax) + 
        aes(x = Year, y = median, group = Petition, color = Petition) +
        geom_line() +
        labs(title = "Tax paid by petition", y = "Shillings", x = "Year") 

```

The analysis of annual data on the website specifically focuses on thse who were
Pro- and Anti- Parris based on petition signatures. As can be seen in the 
previous graph, those who did not sign petititions always had a lower median
tax, but the other two groups were inconsistent.  To address this the site
presents the ratio of the medians of the two group.  A ratio of 1.0 means that
the median tax for the two groups was identical.  A ratio of great than
1 means that the Pro-Parris group had a higher median. A ratio of 
less than one means that the Anti-Parris group had a higher median.

```{r fig.width=5}

ratios <- yearly_tax %>% filter(Petition %in% c("Pro-P", "Anti-P")) %>%
                   select(Year, Petition, median) %>% 
                   pivot_wider(id_cols = c(Year), values_from = median, 
                               names_from = Petition) %>% 
                    mutate(median_ratio = `Pro-P`/`Anti-P`)
ratios %>% kable()
ggplot(ratios, 
        aes(x = Year, y = median_ratio)) +
        geom_point() +
        geom_line(group = 1) +
        geom_hline(yintercept = 1, color = "blue") +
        labs(title = "Ratio of Pro- to Anti- Parris median tax",
             y = "Ratio", x = "Year") 

```

The same analysis can be done but substituting mean for median.

```{r}
ratios <- yearly_tax %>% filter(Petition %in% c("Pro-P", "Anti-P")) %>%
                   select(Year, Petition, mean) %>% 
                   pivot_wider(id_cols = c(Year), values_from = mean, 
                               names_from = Petition) %>% 
                    mutate(mean_ratio = `Pro-P`/`Anti-P`)
ratios %>% kable()
ggplot(ratios, 
        aes(x = Year, y = mean_ratio)) +
        geom_point() +
        geom_line(group = 1) +
        geom_hline(yintercept = 1, color = "blue") +
        labs(title = "Ratio of Pro- to Anti- Parris mean tax",
             y = "Ratio", x = "Year") 
```



The site also looks at individual trajectories using ranks and percentiles 
within each year, allowing analysis of whether an individual's financial
position changed substantially over time. Individuals whose percentile changed
by more than 10 are considered to have moved substantially.

There are many ways to do this, and the method chosen will depend on your 
purposes. The example below is not necessarily the best or most efficient. 
To do the analysis we need to obtain the rank and percentile for each
of the tax amounts within the years.  Finally, we need to compare those. 
In this example we will focus on the 
[1681-1690 comparison shown](https://www.tulane.edu/~salem/1681-90%20Rank.html) 
while providing a framework for the others.

First create a wide data set and calculate ranks and cumulative proportions. 
This allows viewing and sorting of individuals for specific years.

```{r}
persisters <- tax_comparison %>% pivot_wider(id_cols = c(Name, Petition),
                                             names_from = Year,
                                             names_prefix = "X",
                                             values_from = Tax
                                             ) %>%
                                 relocate(X1681, .before = X1690) 

```
Next we want to compare our two selected years and determine the number who
went up, down and stayed the same and how that relates to the petitions.
This analysis is based on the empirical cumulative relative frequency.

To reproduce the analysis on the website we use the dplyr min_rank function
(which is equivalent to rank(ties.method = "min") in base R).  This ranks
all of the tax payments with gaps between the ties.  So if two people 
are tied for the second rank, the person after them would be ranked fourth.
Turn the rank into a percentile by subtracting 1 and dividing by the number 
of tax payers in that year. 


```{r}

comparison <- persisters %>% 
                    mutate(across(starts_with("X"), 
                                               list(
                                                    rank = min_rank
                                                    )
                                        )) %>%
                   filter(!is.na(X1690) & !is.na(X1681)) %>%
                   mutate(
                         X1690_ptile = 100 * (X1690_rank - 1)/100,
                         X1681_ptile = 100* (X1681_rank - 1)/94,
                         compare_1681_1690 =
                         case_when(
                          X1690_ptile -X1681_ptile >=  10 ~ "Up",
                          X1690_ptile -X1681_ptile <= -10 ~ "Down",
                          TRUE ~ "No Change"
                        )  
                    ) 

table(comparison$compare_1681_1690, comparison$Petition, 
      dnn = c("Change", "Petition")) %>%
      kable(caption = "Change in position 1681-1690")


```



## Conclusion

This has illustrated how to reproduce the analyses from the Salem Witchcraft 
Website. The raw data is available at the websie and at 
[the respository for this package](https://github.com/elinw/salem/tree/master/rawdata).
The code used to transform the original data into the files presented here is
also available at that location.

